---
title: "GSEA: filtered vs unfiltered gene list — Human ASM"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    css: assets/report.css
---

<!-- ----------------------------------------------------------------------- -->
<!-- PURPOSE: test whether running GSEA on the FULLY UNFILTERED gene list      -->
<!-- (before low-count pre-filtering) recovers pathway signal that the        -->
<!-- filtered list did not. Self-contained: rebuilds the DDS from the count    -->
<!-- matrix so no re-run of the main pipeline is needed.                       -->
<!-- ----------------------------------------------------------------------- -->

# GSEA: filtered vs unfiltered gene list

The main analysis pre-filters low-count genes before DESeq2. GSEA was then run on
that filtered list (~14,080 genes) and found almost no coordinated pathway signal.
Here we test the alternative: run DESeq2 and GSEA on the **fully unfiltered** gene
list, and compare.

```{r libs, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(fgsea)
library(msigdbr)
library(knitr)
library(kableExtra)
```

## Rebuild the count matrix and metadata (mirrors script 01)

```{r load, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
git_root <- system("git rev-parse --show-toplevel", intern = TRUE)

samples_info <- read.table(
  file.path(git_root, "data", "data-02-Homo_sapiens", "metadata", "metadata.tsv"),
  header = TRUE, sep = "\t", check.names = TRUE
)
samples_info$group <- factor(samples_info$group, levels = c("control", "treatment"))
samples_info$donor <- factor(samples_info$donor)

count_x <- readRDS(
  file.path(git_root, "data", "data-02-Homo_sapiens", "hasapiens", "star_salmon",
            "salmon.merged.gene.SummarizedExperiment.rds")
)

count_genes <- assay(count_x, assayNames(count_x)[1])
gene_symbols <- rowData(count_x)$gene_name
gene_ids     <- rowData(count_x)$gene_id
gene_symbols_saved <- ifelse(
  !is.na(gene_symbols) & nchar(gene_symbols) > 0,
  make.unique(as.character(gene_symbols)),
  make.unique(as.character(gene_ids))
)

count_genes <- as.matrix(count_genes)
mode(count_genes) <- "numeric"
count_genes <- round(count_genes)
storage.mode(count_genes) <- "integer"
rownames(count_genes) <- gene_symbols_saved

# subset to the 8 dex/untreated samples
keep_samples <- intersect(samples_info$sample, colnames(count_genes))
count_genes  <- count_genes[, keep_samples, drop = FALSE]
samples_info <- samples_info[match(colnames(count_genes), samples_info$sample), ]
samples_info$condition <- relevel(factor(samples_info$group), ref = "control")

cat("Genes in matrix (unfiltered):", nrow(count_genes), "\n")
```

## Build DDS and run DESeq2 — WITHOUT pre-filtering

We keep every gene that has a nonzero total count (DESeq2 drops all-zero rows
internally anyway). No `rowSums(counts >= 10)` pre-filter.

```{r deseq-unfiltered, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
dds_unf <- DESeqDataSetFromMatrix(
  countData = count_genes,
  colData   = samples_info,
  design    = ~ donor + condition
)

# Only remove genes that are all-zero (unavoidable), NOTHING else
dds_unf <- dds_unf[rowSums(counts(dds_unf)) > 0, ]
cat("Genes kept (nonzero only):", nrow(dds_unf), "\n")

dds_unf <- DESeq(dds_unf)

res_unf <- results(dds_unf,
                   contrast = c("condition", "treatment", "control"),
                   alpha = 0.05) %>%
  as.data.frame() %>%
  rownames_to_column("gene")
```

## Build the ranked list from the unfiltered results

```{r rank-unfiltered, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ranked_unf <- res_unf %>%
  filter(!is.na(pvalue), !is.na(log2FoldChange), pvalue > 0) %>%
  mutate(rank_metric = sign(log2FoldChange) * -log10(pvalue)) %>%
  arrange(desc(rank_metric)) %>%
  distinct(gene, .keep_all = TRUE)

rank_unf <- ranked_unf$rank_metric
names(rank_unf) <- ranked_unf$gene

cat("Genes in UNFILTERED ranked list:", length(rank_unf), "\n")
```

## Also load the FILTERED ranked list (from the main pipeline) for comparison

```{r rank-filtered, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
res_filt <- read.table(
  file.path(git_root, "results", "human", "DE_treatment_vs_control_all.tsv"),
  header = TRUE, sep = "\t", check.names = TRUE
)
ranked_filt <- res_filt %>%
  filter(!is.na(pvalue), !is.na(log2FoldChange), pvalue > 0) %>%
  mutate(rank_metric = sign(log2FoldChange) * -log10(pvalue)) %>%
  arrange(desc(rank_metric)) %>%
  distinct(gene, .keep_all = TRUE)
rank_filt <- ranked_filt$rank_metric
names(rank_filt) <- ranked_filt$gene

cat("Genes in FILTERED ranked list:", length(rank_filt), "\n")
```

## Fetch gene sets (Hallmark + KEGG + Reactome)

```{r sets, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
get_sets <- function(collection, subcollection = NULL) {
  df <- tryCatch(
    if (is.null(subcollection)) msigdbr(species = "Homo sapiens", collection = collection)
    else msigdbr(species = "Homo sapiens", collection = collection, subcollection = subcollection),
    error = function(e)
      if (is.null(subcollection)) msigdbr(species = "Homo sapiens", category = collection)
      else msigdbr(species = "Homo sapiens", category = collection, subcategory = subcollection)
  )
  split(x = df$gene_symbol, f = df$gs_name)
}

avail <- msigdbr_collections()
subcol_names <- if ("gs_subcollection" %in% names(avail)) avail$gs_subcollection else avail$gs_subcat
kegg_subcol <- if ("CP:KEGG_LEGACY" %in% subcol_names) "CP:KEGG_LEGACY" else "CP:KEGG"

sets <- c(
  get_sets("H"),
  get_sets("C2", kegg_subcol),
  get_sets("C2", "CP:REACTOME")
)
cat("Total gene sets:", length(sets), "\n")
```

## Run GSEA on both ranked lists

```{r gsea, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
set.seed(42)

run_gsea <- function(rv) {
  fgseaMultilevel(pathways = sets, stats = rv, minSize = 15, maxSize = 1000) %>%
    as.data.frame()
}

g_filt <- run_gsea(rank_filt)
g_unf  <- run_gsea(rank_unf)
```

## Head-to-head comparison

```{r compare, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
relevant_kw <- "cytokine|inflamm|immune|interleukin|IL[0-9]|TNF|NFKB|chemokine|glucocortic|steroid|hormone|JAK|STAT|MAPK"

summ <- function(name, res) {
  sig <- res %>% filter(padj < 0.05)
  data.frame(
    List            = name,
    Genes_ranked    = NA_integer_,
    Sig_padj_0.05   = nrow(sig),
    Relevant_sig    = sum(grepl(relevant_kw, sig$pathway, ignore.case = TRUE)),
    Top_by_padj     = if (nrow(sig) > 0) sig %>% arrange(padj) %>% slice(1) %>% pull(pathway) else NA_character_
  )
}

comp <- bind_rows(
  summ("Filtered (~14k genes)",  g_filt),
  summ("Unfiltered (all genes)", g_unf)
)
comp$Genes_ranked <- c(length(rank_filt), length(rank_unf))

comp %>%
  kable(caption = "GSEA on filtered vs unfiltered ranked list") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Top hits — unfiltered list (nominal pval < 0.05, for inspection)

Even if nothing survives padj, the top NES by nominal p-value shows whether the
biology is *directionally* sensible.

```{r top-unf, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
g_unf %>%
  filter(pval < 0.05) %>%
  arrange(pval) %>%
  mutate(NES = round(NES, 2), pval = signif(pval, 3), padj = signif(padj, 3)) %>%
  select(pathway, NES, pval, padj, size) %>%
  head(20) %>%
  kable(caption = "Unfiltered GSEA — top sets by nominal p-value") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

<!-- --------------------------------------------------------------------- -->
<!-- After running: paste the two tables back. If unfiltered is no better   -->
<!-- (expected), we keep the filtered list and make ORA the main result.    -->
<!-- --------------------------------------------------------------------- -->
