---
title: "27200 Data Driven Bioengineering - Transcriptomics"
subtitle: "Workshop material"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook
---
# Welcome
 
```{r,echo=FALSE,include=FALSE}
#  variables
rep_org <- "Data Science Platform" 
rep_core  <- "Juliana Assis"
rep_core_email  <- "jasge@dtu.dk"
rep_core2 <- "Sebastian Schulz"
rep_core_email2 <- "sebschu@dtu.dk"
rep_core3 <- "Alberto Pallejà Caro"
rep_core_email3 <- "apca@dtu.dk"
rep_lecturer       <- "Lasse Ebdrup Pedersen"
rep_lecturer_email <- "laeb@dtu.dk"
rep_lecturer_title <- "Senior Researcher, DTU Bioengineering"

# Combine all responsibles into a single string
rep_all <- paste0(
  rep_core, " (", rep_core_email, ")<br>",
  rep_core2, " (", rep_core_email2, ")<br>",
  rep_core3, " (", rep_core_email3, ")"
)

rep_lecturer_full <- paste0(rep_lecturer, " (", rep_lecturer_email, ") — ", rep_lecturer_title)
```

<br>

::: boxy
__Organisation:__ `r rep_org`
__Lecturer (theory):__ `r rep_lecturer_full`
__Responsibles:__ `r paste0(rep_all," ")`
:::

<br>


**Welcome to the Transcriptomics week of 27200 Data-driven Bioengineering!**

This session is **Week 6** of the course (lecture: **Thursday, 08 October 2026**), part of the omics block. The theme of the week: **transcript abundance as a dynamic — but incomplete — regulatory readout**.

We cover the theory behind RNA sequencing and the full practical workflow — from raw reads to biological interpretation: the RNA-seq workflow and count matrices, normalization, batch effects revisited, PCA, differential expression, and what transcriptomics reflects and misses.

The material is designed for people with **no or little practical experience** in transcriptomic analysis using RNA sequencing (RNA-seq).

**Objectives**

* Theory on Illumina short-read sequencing technology for transcriptome analysis
* Theory and practical sessions for
   * RNA-seq data processing using the nf-core/rnaseq pipeline
   * Quality control and exploratory analysis (normalization, PCA, sample correlation)
   * Differential expression analysis with DESeq2 — design formulas, shrinkage, multiple testing
   * Enrichment analysis (ORA & GSEA) — gaining biological insights from transcriptomic data
   * Critically assessing what transcript abundance can and cannot tell you about biology

**Datasets**

* 🦠 *Staphylococcus aureus* — biofilm vs planktonic lifestyle over time (strains USA-100 and USA-500)
* 🧑‍🔬 *Homo sapiens* — airway smooth muscle cells treated with dexamethasone (paired donor design)

:::boxy
You can also explore the [27200 Data Driven Bioengineering - Transcriptomics repository](https://github.com/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)
:::

**Draft agenda** (4-hour block: ~2 h lecture + ~2 h guided exercises, in groups of 4)

| Time | Session |
|---|---|
| ☕ **10:00–10:15** | Welcome and setup |
| 🔬 **10:15–11:00** | Theory: Illumina sequencing, RNA-seq workflow & the nf-core/rnaseq pipeline |
| 📊 **11:00–11:45** | Theory: count matrices, normalization, PCA & differential expression (+ quizzes) |
| ⏸️ **11:45–12:00** | Break |
| 💻 **12:00–12:55** | Exercises: *Quality Control & Exploratory Data Analysis* |
| 🧬 **12:55–13:50** | Exercises: *Differential Expression & Functional Enrichment (ORA/GSEA)* |
| 🏁 **13:50–14:00** | Wrap-up: key insight + muddy point |


:::boxy
Material for the workshop is located at:
[dsp_transcriptomics_27200-Data-driven-bioengineering](https://github.com/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)
:::

Below are two setup options for the practical activities:

>Run the workshop in the cloud (no need to install anything). 

**Launch the app:**

[![Open in GitHub Codespaces](https://github.com/codespaces/badge.svg)](https://codespaces.new/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)


>Run the workshop locally on your machine. 

To run it on your own machine, install the following packages:

```{r, eval=FALSE}
# Install BiocManager if needed
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Bioconductor packages
BiocManager::install(c(
  "DESeq2",
  "SummarizedExperiment",
  "apeglm",
  "EnhancedVolcano"
))

# CRAN packages
packages <- c(
  "dplyr",
  "ggplot2",
  "ggrepel",
  "RColorBrewer",
  "rmarkdown",
  "tidyr",
  "plotly",
  "pheatmap",
  "ggpubr",
  "here",
  "mulea"
)

install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org/")
  } else {
    message(pkg, " already installed.")
  }
}

invisible(lapply(packages, install_if_missing))

message("All requested R packages are installed and ready to use!")
```
