---
title: "27200 Data Driven Bioengineering - Transcriptomics"
subtitle: "Workshop material"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook
---
# Welcome
 
```{r,echo=FALSE,include=FALSE}
#  variables
rep_lecturer <- "Lasse Ebdrup Pedersen (laeb@dtu.dk), Senior Researcher - DTU Bioengineering"

rep_all <- paste0(
  "Juliana Assis, Senior Data Scientist - Biotechnology Research Institute for the Green Transition BRIGHT (jasge@dtu.dk)<br>",
  "Sebastian Schulz, Senior Data Scientist - BRIGHT (sebschu@dtu.dk)<br>",
  "Alberto Pallejà Caro, Team Lead Data Science Platform - BRIGHT (apca@dtu.dk)"
)
```

<br>

::: boxy
__Lecturer (theory):__ `r rep_lecturer`<br>
__Responsibles:__ `r rep_all`
:::

<br>


**Welcome to the Transcriptomics week of 27200 Data-driven Bioengineering!**

This session is **Week 6** of the course (lecture: **Thursday, 08 October 2026**), part of the omics block. The theme of the week: **transcript abundance as a dynamic — but incomplete — regulatory readout**.

We cover the theory behind RNA sequencing and the full practical workflow — from raw reads to biological interpretation: the RNA-seq workflow and count matrices, normalization, batch effects revisited, PCA, differential expression, and what transcriptomics reflects and misses.

The material is designed for people with **no or little practical experience** in transcriptomic analysis using RNA sequencing (RNA-seq).

**Objectives**

* Theory on Illumina short-read sequencing technology for transcriptome analysis
* Theory and practical sessions for
   * RNA-seq data processing using the nf-core/rnaseq pipeline
   * Quality control and exploratory analysis (normalization, PCA, sample correlation)
   * Differential expression analysis with DESeq2 — design formulas, shrinkage, multiple testing
   * Enrichment analysis (ORA & GSEA) — gaining biological insights from transcriptomic data
   * Critically assessing what transcript abundance can and cannot tell you about biology

**Datasets**

* 🦠 *Staphylococcus aureus* — biofilm vs planktonic lifestyle over time (strains USA-100 and USA-500)
* 🧑‍🔬 *Homo sapiens* — airway smooth muscle cells treated with dexamethasone (paired donor design)

:::boxy
**Pick ONE dataset** — whichever interests you more, microbial or biomedical — and work through its three notebooks in order (**01 quality control → 02 differential expression → 03 functional enrichment**). Depth beats coverage: it is better to interpret one dataset well than to rush through both. If your group finishes early, start on the other dataset.
:::

:::boxy
You can also explore the [27200 Data Driven Bioengineering - Transcriptomics repository](https://github.com/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)
:::

**Draft agenda** (4-hour block: ~2 h lecture + ~2 h guided exercises, in groups of 4)

| Time | Session |
|---|---|
| ☕ **10:00–10:15** | Welcome and setup |
| 🔬 **10:15–11:00** | Theory: Illumina sequencing, RNA-seq workflow & the nf-core/rnaseq pipeline |
| 📊 **11:00–11:45** | Theory: count matrices, normalization, PCA & differential expression (+ quizzes) |
| ⏸️ **11:45–12:00** | Break |
| 💻 **12:00–12:55** | Exercises: *Quality Control & Exploratory Data Analysis* |
| 🧬 **12:55–13:50** | Exercises: *Differential Expression & Functional Enrichment (ORA/GSEA)* |
| 🏁 **13:50–14:00** | Wrap-up: key insight + muddy point |


:::boxy
Material for the workshop is located at:
[dsp_transcriptomics_27200-Data-driven-bioengineering](https://github.com/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)
:::

Below are two setup options for the practical activities:

>Run the workshop in the cloud (no need to install anything).

<div class="warning">
  <strong>⚠️ Warning:</strong> Change the configuration at the Machine type: increase it to 4-core.
</div>

**Launch the app:**

[![Open in GitHub Codespaces](https://github.com/codespaces/badge.svg)](https://codespaces.new/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)


>Run the workshop locally on your machine. 

To run it on your own machine, install the following packages:

```{r, eval=FALSE}
# ============================================================
# DSP Transcriptomics Workshop – Package Installation Script
# ============================================================
# Run this script ONCE before the workshop to install all
# required R packages. Works on macOS, Linux, and Windows.
# Tested with R >= 4.3.
# ============================================================

## ---- 0. Helper: install only if missing --------------------

install_if_missing <- function(pkg, installer, ...) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Installing: ", pkg)
    installer(pkg, ...)
  } else {
    message("Already installed: ", pkg)
  }
}

## ---- 1. CRAN packages --------------------------------------

cran_pkgs <- c(
  "tidyverse",
  "reshape2",
  "RColorBrewer",
  "pheatmap",
  "heatmaply",
  "factoextra",
  "knitr",
  "kableExtra",
  "DT",
  "plotly",
  "ggpubr",
  "gggenes",     # operon plots (S. aureus)
  "gprofiler2",  # ORA via g:Profiler (human)
  "msigdbr",     # MSigDB gene sets for GSEA (human)
  "remotes"      # needed for GitHub installs below
)

for (pkg in cran_pkgs) {
  install_if_missing(pkg, install.packages,
                     dependencies = TRUE)
}

## ---- 2. Bioconductor packages ------------------------------

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
# Pin to the release that matches R 4.4; BiocManager picks the right
# version automatically, but you can force with: BiocManager::install(version = "3.19")

bioc_pkgs <- c(
  "DESeq2",
  "apeglm",       # recommended shrinkage estimator used with DESeq2
  "fgsea",
  "KEGGREST",
  "EnhancedVolcano",
  "org.EcK12.eg.db"  # E. coli gene annotation (E. coli chapters)
)

for (pkg in bioc_pkgs) {
  install_if_missing(pkg, BiocManager::install,
                     update = FALSE, ask = FALSE)
}

## ---- 3. GitHub packages ------------------------------------

# mulea is not on CRAN/Bioconductor yet
if (!requireNamespace("mulea", quietly = TRUE)) {
  message("Installing mulea from GitHub (ELTEbioinformatics/mulea)")
  remotes::install_github("ELTEbioinformatics/mulea")
} else {
  message("Already installed: mulea")
}

## ---- 4. Verification ---------------------------------------

all_pkgs <- c(
  cran_pkgs,
  bioc_pkgs,
  "mulea"
)
# remove helper-only packages from the check list
check_pkgs <- setdiff(all_pkgs, "remotes")

cat("\n========== Installation check ==========\n")
ok  <- character(0)
nok <- character(0)

for (pkg in check_pkgs) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    ok <- c(ok, pkg)
  } else {
    nok <- c(nok, pkg)
  }
}

cat("OK  (", length(ok),  "):", paste(ok,  collapse = ", "), "\n\n")

if (length(nok) > 0) {
  cat("FAILED (", length(nok), "):", paste(nok, collapse = ", "), "\n")
  cat("Please re-run the script or install the above packages manually.\n")
} else {
  cat("All packages installed successfully. You are ready for the workshop!\n")
}
cat("=========================================\n")
```
