---
title: "27200 Data Driven Bioengineering - Transcriptomics"
subtitle: "Workshop material"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook
---
# Welcome
 
```{r,echo=FALSE,include=FALSE}
# R starts in the C/US-ASCII locale here, which makes every non-ascii character
# in a plot come out as a literal escape (Palleja<U+00E0>, GSEA <U+2014> ...).
# Asking for a UTF-8 locale fixes all of them at once; ragg draws them cleanly.
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))
knitr::opts_chunk$set(dev = "ragg_png")

#  variables
rep_lecturer <- "Lasse Ebdrup Pedersen (laeb@dtu.dk), Senior Researcher - DTU Bioengineering"

rep_all <- paste0(
  "Juliana Assis, Senior Data Scientist - Biotechnology Research Institute for the Green Transition BRIGHT (jasge@dtu.dk)<br>",
  "Sebastian Schulz, Senior Data Scientist - BRIGHT (sebschu@dtu.dk)<br>",
  "Alberto Palleja Caro, Team Lead Data Science Platform - BRIGHT (apca@dtu.dk)"
)
```

<br>

::: boxy
__Lecturer (theory):__ `r rep_lecturer`<br>
__Responsibles:__ `r rep_all`
:::

<br>


**Welcome to the Transcriptomics week of 27200 Data-driven Bioengineering!**

This session is **Week 6** of the course (lecture: **Thursday, 08 October 2026**), part of the omics block. The theme of the week: **transcript abundance as a dynamic - but incomplete - regulatory readout**.

We cover the theory behind RNA sequencing and the full practical workflow - from raw reads to biological interpretation: the RNA-seq workflow and count matrices, normalization, batch effects revisited, PCA, differential expression, and what transcriptomics reflects and misses.

The material is designed for people with **no or little practical experience** in transcriptomic analysis using RNA sequencing (RNA-seq).

**Objectives**

* Theory on Illumina short-read sequencing technology for transcriptome analysis
* Theory and practical sessions for
   * RNA-seq data processing using the nf-core/rnaseq pipeline
   * Quality control and exploratory analysis (normalization, PCA, sample correlation)
   * Differential expression analysis with DESeq2 - design formulas, shrinkage, multiple testing
   * Enrichment analysis (GSEA, plus ORA in the advanced bacterial track) - gaining biological insights from transcriptomic data
   * Critically assessing what transcript abundance can and cannot tell you about biology

**Datasets**

* 🧑‍🔬 *Homo sapiens* - airway smooth muscle cells treated with dexamethasone (paired donor design). This is the one we work on together.
* 🦠 *Staphylococcus aureus* - biofilm vs planktonic over time, strains USA-100 and USA-500. Advanced material, if you want to go further.

:::boxy
**Work through the human dataset**: the three notebooks in order (**01 quality control → 02 differential expression → 03 functional enrichment**). Each one saves results that the next one reads, so the order matters. Depth beats coverage: it is better to interpret one dataset well than to rush. If you finish early, the *S. aureus* material takes the same ideas further, with a time course and two strains.
:::

:::boxy
You can also explore the [27200 Data Driven Bioengineering - Transcriptomics repository](https://github.com/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)
:::

**Draft agenda** (4-hour block: lecture + guided exercises, in groups of 4)

| Time | Session |
|---|---|
| ☕ **8:00–8:15** | Welcome and setup. **Start the pipeline run now** - it takes about 25 minutes and finishes while we talk |
| 🔬 **8:15–9:00** | Theory: Illumina sequencing, RNA-seq workflow & the nf-core/rnaseq pipeline |
| 📊 **9:00–9:45** | Theory: count matrices, normalization, PCA & differential expression (+ quizzes) |
| ⏸️ **9:45–10:00** | Break |
| 💻 **10:00–10:55** | Exercises: *Quality Control & Exploratory Data Analysis* |
| 🧬 **10:55–11:50** | Exercises: *Differential Expression & Functional Enrichment (GSEA)* |
| 🏁 **11:50–12:00** | Wrap-up: key insight + muddy point |


:::boxy
Material for the workshop is located at:
[dsp_transcriptomics_27200-Data-driven-bioengineering](https://github.com/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)
:::

**How to run the practicals.** Use **GitHub Codespaces** - that is what we use in class,
it needs nothing installed, and everyone gets an identical environment. The other two
options below are for people who specifically want them, and are not supported during
the session.

### Option 1: Codespaces (recommended)

<div class="warning">
  <strong>⚠️ Before you start:</strong> change the Machine type to <strong>4-core</strong> when you create the Codespace.
</div>

[![Open in GitHub Codespaces](https://github.com/codespaces/badge.svg)](https://codespaces.new/biosustain/dsp_transcriptomics_27200-Data-driven-bioengineering)

When it has started, open `02_notebooks/` and begin with the human dataset.

### Option 2: Docker

The Codespace runs from an image you can also start yourself. It already contains R, all
the packages, Jupyter with the R kernel, and Nextflow. From the folder where you cloned
the repository:

```bash
docker run --rm -it -p 8888:8888 -v "$PWD":/work -w /work \
  ghcr.io/biosustain/dsp_transcriptomics_27200-data-driven-bioengineering:1.1.0 \
  jupyter lab --ip=0.0.0.0 --port=8888 --no-browser --allow-root
```

Open the `http://127.0.0.1:8888/lab?token=...` link it prints, then go to `02_notebooks/`.
On an Apple Silicon Mac add `--platform linux/amd64`.

### Option 3: your own R installation

Only sensible if you are comfortable with R and with sorting out package installations
yourself, Bioconductor ones included. If it breaks on the morning of the class, switch to
a Codespace rather than lose the session to it. Install the packages well in advance:

```{r, eval=FALSE}
# ============================================================
# DSP Transcriptomics Workshop – Package Installation Script
# ============================================================
# Run this script ONCE before the workshop to install all
# required R packages. Works on macOS, Linux, and Windows.
# Tested with R >= 4.3.
# ============================================================

## ---- 0. Helper: install only if missing --------------------

install_if_missing <- function(pkg, installer, ...) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Installing: ", pkg)
    installer(pkg, ...)
  } else {
    message("Already installed: ", pkg)
  }
}

## ---- 1. CRAN packages --------------------------------------

cran_pkgs <- c(
  "tidyverse",
  "reshape2",
  "RColorBrewer",
  "pheatmap",
  "knitr",
  "kableExtra",
  "DT",
  "ggpubr",
  "gggenes",     # operon plots (S. aureus)
  "msigdbr",     # MSigDB gene sets for GSEA (human)
  "ragg",        # draws plots with proper unicode
  "remotes"      # needed for GitHub installs below
)

for (pkg in cran_pkgs) {
  install_if_missing(pkg, install.packages,
                     dependencies = TRUE)
}

## ---- 2. Bioconductor packages ------------------------------

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
# Pin to the release that matches R 4.4; BiocManager picks the right
# version automatically, but you can force with: BiocManager::install(version = "3.19")

bioc_pkgs <- c(
  "DESeq2",
  "apeglm",       # recommended shrinkage estimator used with DESeq2
  "fgsea",
  "EnhancedVolcano"
)

for (pkg in bioc_pkgs) {
  install_if_missing(pkg, BiocManager::install,
                     update = FALSE, ask = FALSE)
}

## ---- 3. GitHub packages ------------------------------------

# mulea is not on CRAN/Bioconductor yet
if (!requireNamespace("mulea", quietly = TRUE)) {
  message("Installing mulea from GitHub (ELTEbioinformatics/mulea)")
  remotes::install_github("ELTEbioinformatics/mulea")
} else {
  message("Already installed: mulea")
}

## ---- 4. Verification ---------------------------------------

all_pkgs <- c(
  cran_pkgs,
  bioc_pkgs,
  "mulea"
)
# remove helper-only packages from the check list
check_pkgs <- setdiff(all_pkgs, "remotes")

cat("\n========== Installation check ==========\n")
ok  <- character(0)
nok <- character(0)

for (pkg in check_pkgs) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    ok <- c(ok, pkg)
  } else {
    nok <- c(nok, pkg)
  }
}

cat("OK  (", length(ok),  "):", paste(ok,  collapse = ", "), "\n\n")

if (length(nok) > 0) {
  cat("FAILED (", length(nok), "):", paste(nok, collapse = ", "), "\n")
  cat("Please re-run the script or install the above packages manually.\n")
} else {
  cat("All packages installed successfully. You are ready for the workshop!\n")
}
cat("=========================================\n")
```
